The Multi-Stencil Language

### 2016-01-21 ARTICLES/IJPP/article/formalism.tex
article IJPP Helene 21/01/2016

\noindent \textbf{Definitions}
%\begin{mydef}
\begin{itemize}
\item A computation domain $D$ is a subpart  a mesh entities group, $D \subset E \in \mathcal{E}$.
\item The set of computation domains of a numerical simulation is denoted $\mathcal{D}$.
\item A neighborhood $n$ is a function which for a given entity $e \in E_i$, returns a set of $m$ entities in $E_j$, $n : E_i \rightarrow E_j^m$. One can notice that $i = j$ is possible.
\item The set of neihborhood functions in a numerical simulation is denoted $\mathcal{N}$.
\item A single data is written during a computation and is denoted $w \in \Delta$. Thus, the computation domain of the computation is $D \subset E_w=entity(w)$.
\item A set of data to read and use in a computation is denoted $R$. Each element of $R$ is a tuple $(r,n)$, where $r \in \Delta$ and $n$ is a neighborhood function such that $n : E_w \rightarrow entity(r)^m$. The neighborhood indicates which entities, on which $r$ is applied, will be read during the computation.
\item A numerical expression $\text{exp}_{w}: entity(w) \times R^n \rightarrow V_{w}$ is a function which computes the values of the written quantity $w \in \Delta$ for a given entity $e \in entity(w)=E_w$, using a set of input data $R$.
\item A computation kernel $k$ of a numerical simulation is defined as $c(R,w,\text{exp},D)$, where $R$ is the set of data read, $w \in \Delta$ is the unique data written, $\text{exp}$ is a numerical expression, and $D \in \mathcal{D}$ is the computation domain.
\end{itemize}
%\end{mydef}
%If the number of computations in $\Gamma$ is $card(\Gamma)=n$, such that $\bigcup_{i=0}^{n-1}c_i = \Gamma$, then $\bigcup_{i=0}^{m-1}R_i \cup w_i \subseteq \Delta$.

Most of the time we can dissociate two types of kernel computations.

\noindent \textbf{Definitions}
%\begin{mydef}
\begin{itemize}
\item We denote by $identity$ the identity function $x=x$.
\item A kernel computation $k(R,w,\text{exp},D) \in \Gamma$ is a stencil kernel $\iff \exists (r,n) \in R$ such that $n \neq identity$.
\item A kernel computation $k(R,w,\text{exp},D) \in \Gamma$ is a local kernel $\iff \forall (r,n) \in R$, $n = identity$.
\end{itemize}

@@

A reduction computation is a computation which takes data applied onto meshes and returns a data with a dimension reduced to zero, in other words a scalar. A reduction is typically used to compute the convergence criteria of the time loop of the simulation. Occasionally reductions can also be performed during a time iteration. %This last usage is typically done to have a conditionnal branch to choose one computation or another, or to compute a dynamic time step for example.

\medskip
\noindent \textbf{Definitions}
%\begin{mydef}
\begin{itemize}
\item A reduction computation $r$ of a numerical simulation is defined as $r(R,w,red)$, where $R$ is the set of data read, $w \in \mathcal{S}$ is the scalar written, and $\text{red}$ is an associative reduction function.
% \item For any $i,j$ such that $r_i \in R$ and $r_j \in R$, if $dim(mesh(r_i))=dim(mesh(r_j))=n$.
% \item The data written by the reduction $w$ verifies $dim(mesh(w))<n$.
\item A reduction function $\text{red}: \Delta^n \rightarrow \mathcal{S}$ is an operation which reduces a set of data in $\Delta$ (on it sentities) to a single scalar variable in $\mathcal{S}$. The reduction function must be associative.
\item The set of $n$ ordered computations (kernel or reduction) of a numerical simulation is denoted $\Gamma = [c_i]_{0 \leq i \leq n-1}$, such that $\forall c_i,c_j \in \Gamma$, if $i \leq j$, then $c_i$ is computed before $c_j$.
\item Finally, a \textit{multi-stencil program} is defined by the septuplet $\mathcal{MSP}(T,\mathcal{M},\mathcal{E},\mathcal{D},\Delta, \mathcal{S},\Gamma)$.
\end{itemize}
%!!!!!!!  est-ce qu'on fusionne le maillage, les entites et les domaines ?????
%\end{mydef}

>>>>

\noindent \textbf{Definitions}
%\begin{mydef}
\begin{itemize}
\item A computation domain $D$ is a subpart of a mesh entities group, $D \subset E \in \mathcal{E}$.
\item The set of computation domains of a numerical simulation is denoted $\mathcal{D}$.
\item A neighborhood $n$ is a function which for a given entity $e \in E_i$, returns a set of $m$ entities in $E_j$, $n : E_i \rightarrow E_j^m$. One can notice that $i = j$ is possible. Most of the time, such a neighborhood is called a stencil shape.
\item The set of neihborhood functions in a numerical simulation is denoted $\mathcal{N}$.
\end{itemize}

\begin{mydef}
A computation kernel $k$ of a numerical simulation is defined as $k(R,(w,D),comp)$, where 
\begin{itemize}
\item $(w,D) \in \Delta \times \mathcal{D}$ is the single data modified by the computation kernel onto a given computation domain. Thus, $w \in \Delta$ and $D \in \mathcal{D}$ is the computation domain on which $w$ is computed, $D \subset E_w=entity(w)$.
\item $R$ is the set of tuples $(r,n)$, where $r \in \Delta$ and $n \in \mathcal{N}$ is a neighborhood function such that $n : E_w \rightarrow entity(r)^m$. The neighborhood indicates which entities of $r$ will be read in order to compute a single entity of $w$. 
\item $comp$ is the numerical computation which returns a value from a set of input values, $comp: V^n \rightarrow V$.
\end{itemize}
% \item A numerical expression $\text{exp}_{w}: entity(w) \times R^n \rightarrow V_{w}$ is a function which computes the values of the written quantity $w \in \Delta$ for a given entity $e \in entity(w)=E_w$, using a set of input data $R$.
% \item A computation kernel $k$ of a numerical simulation is defined as $c(R,w,\text{exp},D)$, where $R$ is the set of data read, $w \in \Delta$ is the unique data written, $\text{exp}$ is a numerical expression, and $D \in \mathcal{D}$ is the computation domain.
% \end{itemize}
%$R$ is the set of data read, $w \in \Delta$ is the unique data written, $\text{exp}$ is a numerical expression, and $D \in \mathcal{D}$ is the computation domain.
\end{mydef}

$comp$ represents the actual numerical expression which is computed by a kernel. We can dissociate three types of kernel computations.

\noindent \textbf{Definitions}
%\begin{mydef}
\begin{itemize}
\item We denote by $identity$ the identity function $x=x$.
\item A kernel computation $k(R,(w,D),comp)$ is a \emph{stencil kernel} $\iff \exists (r,n) \in R$ such that $n \neq identity$.
\item A kernel computation $k(R,(w,D),comp)$ is a \emph{local kernel} $\iff \forall (r,n) \in R$, $n = identity$.
\item A kernel computation $k(R,(w,D),comp)$ is a \emph{reduction kernel} if $w$ is a scalar, in other words $\iff dim(mesh(entity(w)))=0$.
\end{itemize}

@@

\medskip
A reduction computation is a computation which takes data applied onto meshes and returns a data with a dimension reduced to zero, in other words a scalar. A reduction is typically used to compute the convergence criteria of the time loop of the simulation. Occasionally reductions can also be performed during a time iteration. %This last usage is typically done to have a conditionnal branch to choose one computation or another, or to compute a dynamic time step for example.

\noindent \textbf{Property}
Considering a reduction kernel $k(R,(w,D),comp)$, $comp$ must be a binary and associative operation on the type $V$, $comp: V \times V \rightarrow V$.

\noindent \textbf{Definitions}
\begin{itemize}
\item The set of $n$ ordered computation kernels of a numerical simulation is denoted $\Gamma = [k_i]_{0 \leq i \leq n-1}$, such that $\forall k_i,k_j \in \Gamma$, if $i \leq j$, then $k_i$ is computed before $k_j$.
\item Finally, a \textit{multi-stencil program} is defined by the septuplet 
\begin{equation*}
\mathcal{MSP}(T,\mathcal{M},\mathcal{E},\mathcal{D},\Delta, \mathcal{S},\Gamma)
\end{equation*}
\end{itemize}

### 2016-01-21 ARTICLES/IJPP/article/parallelism.tex
article IJPP Helene 21/01/2016

Multi-stencil mesh-based numerical simulation can be parallelized in various ways and is an interesting kind of application to take advantage of modern heterogeneous HPC architectures, mixing clusters, multi-cores CPUs, vectorization units, GPGPU and many-core accelerators.

@@

\begin{mydef}
\textit{
For $n$ the number of computations in $\Gamma$, and for $i,j$ such that $i<j<n$, a \textit{synchronisation} is needed between $c_i$ and $c_j$, denoted $c_i \pprec c_j$, if $\exists (r,n) \in R_j$ with $n\neq identity$ ($c_j$ is a stencil computation), and such that $\{w_i\} \subset R_j$. Moreover, the data to update is $\{w_i\} \cap R_j = \{w_i$\}.}
\end{mydef}

@@

\begin{mydef}
\textit{A synchronization between two computations $c_i \pprec c_j$ is defined as a specific computation $sync(R,w,exp,D)$ where $R=\{w_i\}$, $w=w_i$, $exp=identity$ and $D=\emptyset$.}
\end{mydef}

\begin{mydef}
If a synchronization is needed before $c_j$ (on the data $w_i$), $c_j$ is replaced in $\Gamma_{sync}$ by the list
\begin{equation*}
\Gamma_{sync} = [sync(\{w_i\},w_i,identity,\emptyset), c_j]
\end{equation*}
\end{mydef}

\begin{mydef}
The concatenation of two ordered lists of respectively $n$ and $m$ computations $l_1=[c_i]_{0 \leq i \leq n-1}$ and $l_2=[c'_i]_{0 \leq i \leq m-1}$ is denoted $l_1 \cdot l_2$ and is equal to a new ordered list $l_3=[c_0,\dots,c_{n-1},c'_0,\dots,c'_{m-1}]$.
\end{mydef}

\begin{mydef}
From the ordered list of computation $\Gamma$, a new ordered list $\Gamma_{sync}$ is obtained from the call $\Gamma_{sync} = F_{sync}(\Gamma,0)$, where $F_{sync}$ is the recursive function defined as
\begin{equation*}
F_{sync}(\Gamma,j) = 
\begin{cases} 	[\Gamma[j]] \cdot F_{sync}(\Gamma,j+1) & \mbox{if }\forall i<j \mbox{, } c_i \not\pprec c_j\\
				[sync(\{w_i\},w_i,identity,\emptyset), c_j] \cdot F_{sync}(\Gamma,j+1) & \forall i<j \text{, and }c_i \pprec c_j\}\\
				[] & \mbox{if }j=|\Gamma|.
\end{cases}
\end{equation*}
\end{mydef}

@@

% \begin{myprop}
% Denoting four computations $c_l$, $c_k$, $c_i$ and $c_j$ of $\Gamma$, with $l<k$, $i<j$, and where $c_l \pprec c_k$ and $c_i \pprec c_j$. Then, computations $c_k$ and $c_j$ are tansformed to $c^*_k(c_k,update(w_l))$ and $c^*_j(c_j,update(w_i))$. If $i<k$, a single update call can be performed at $k$ such that $c^*_k(c_k,update(w_l\cup w_i))$, and in this case, the computation $c_j$ is not transformed.
% \end{myprop}

% \begin{proof}
% By definition, the synchronization of $w_i$ can be performed between $c_i$ and $c_j$. As $i<k$ the computation $c_i$ is not performed between $k$ and $j$. As a result, the synchronization of $w_i$ can be performed between $c_i$ and $c_k$.
% \end{proof}

% In other words, for each synchronization in $\Gamma_{data}$, if the computation $c_i$ from which $c_j$ is synchronized occurs before the last synchronization, the two synchronizations can be reduced to one with the union of the data to update.

@@

\begin{mydef}
For two computations $c_i(R_i,w_i,\text{exp}_i,D_i)$ and $c_j(R_j,w_j,\text{exp}_j,D_j)$, with $i < j$, it is said that $c_j$ is data dependant from $c_i$, denoted $c_i\prec c_j$, if $\{w_i\} \cap R_j \neq \emptyset$. In this case, $c_i$ has to be computed before $c_j$. The binary relation $c_i\prec c_j$ represents a \textit{dependency}.
\end{mydef}

@@

% \begin{mydef}
% For two computations $c_i(R_i,w_i,\text{exp}_i)$ and $c_j(R_j,w_j,\text{exp}_j)$, a time dependency is denoted $c_i \blacktriangleleft c_j$ and means that $c_i$ has to be computed before $c_j$.
% \end{mydef}

@@

% \begin{myprop}
% For two computations $c_i(R_i,w_i,\text{exp}_i)$ and $c_j(R_j,w_j,\text{exp}_j)$ such that $c_i \prec c_j$, $c_i \blacktriangleleft c_j$ is verified.
% \end{myprop}

% \begin{proof}
% By definition $c_i \prec c_j$ means that $\{w_i\}\cap R_j \neq \emptyset$ and that $c_i$ has to computed before $c_j$. %Thus, the relation $c_i \prec c_j$ is stronger than $c_i \blacktriangleleft c_j$.
% \end{proof}

@@

From an ordered list of computations $\Gamma_{sync}$, a directed dependency graph $\Gamma_{dep}(V,A)$ can be built finding all pairs of computations $c_i(R_i,w_i,\text{exp}_i,D_i)$ and $c_j(R_j,w_j,\text{exp}_j,D_j)$, with $i<j$, such that $c_i \prec c_j$. %Those time dependencies can be found from pairs of data dependencies $c_i \prec c_j$.

@@

\begin{mydef}
From the ordered list $\Gamma_{sync}$ of computations $c(R,w,\text{exp},D)$, a directed dependency graph $\Gamma_{dep}(V,A)$ is obtained from the call $T_{dep}(\Gamma_{sync},0)$, where $F_{dep}$ is the recursive function
\begin{equation*}
F_{dep}(\Gamma_{sync},j) = 
\begin{cases} 	(\{\},\{\}) & \mbox{if }j=|\Gamma_{sync}|\\
				(c_j, \{(\overset{\frown}{c_i,c_j})\mbox{, }\forall i < j \mbox{, } c_i\prec c_j \})\cup F_{dep}(\Gamma_{sync},j+1) & \mbox{if }j<|\Gamma_{sync}|
\end{cases}
\end{equation*}
\end{mydef}

This constructive function is possible because the input is an ordered list. Actually, if $c_i\prec c_j$ then $i<j$. As a result, $c_i$ is already in $V$ when the arc $(\overset{\frown}{c_i,c_j})$ is built.

@@

As a result, the hybrid parallelization of a multi-stencil program $\mathcal{MSP}$, produces $k$ multi-stencil programs
\begin{equation*}
\mathcal{MSP}_k(T,\mathcal{M}_k,\mathcal{E},\mathcal{D},\Delta_k,\Gamma_{dep}).
\end{equation*}
The set of computations $\Gamma_{dep}$ is a dependency graph between computations $c_i$ of $\Gamma$ and the synchronization computations $sync_i$ added into $\Gamma_{sync}$. $\Gamma_{dep}$ can be built from the call to 
\begin{equation*}
F_{dep}(F_{sync}(\Gamma,0),0).
\end{equation*}

% Using the function $T_{task}$ to build $\Gamma_{task}$, however, duplication of dependencies may occur because of the transitivity of the relation $\blacktriangleleft$. Actually, as the relation $\prec$ verifies the relation $\blacktriangleleft$, and as $\blacktriangleleft$ is transitive, if for three computations $c_k(R_k,w_k,\text{exp}_k)$, $c_i(R_i,w_i,\text{exp}_i)$ and $c_j(R_j,w_j,\text{exp}_j)$, with $k<i<j$ and where $c_k \prec c_i$, $c_i \prec c_j$ and $c_k \prec c_j$, then the transitivity $c_k \blacktriangleleft c_j$ is verified because of the relations $c_k \blacktriangleleft c_i$ and $c_i \blacktriangleleft c_j$. However, $c_k \blacktriangleleft c_j$ is also directly represented by the binary relation $c_k \prec c_j$. As a result, a duplication of the dependency is created in $\Gamma_{task}$. In Figure~\ref{fig:duplication} an example of duplication is given for the relation $c_1 \blacktriangleleft c_4$.

@@

% Another view of the relations $\prec$ and $\blacktriangleleft$ is that $c_i\blacktriangleleft c_j$ in $\Gamma_{task}$ is a path from $c_i$ to $c_j$ of any size, while $c_i \prec c_j$ is a path of size $1$ from $c_i$ to $c_j$. And in any case a path in the graph represents a time dependency between computations. Scheduling a graph dependency which do not contains duplication of information is easier. 
% %For this reason, we define a reduction to apply to $\Gamma_{task}$, and which is performed as a post-transformation. In the example of Figure~\ref{fig:duplication} the dependency represented by the path $c_1 \blacktriangleleft c_2 \blacktriangleleft c_3 \blacktriangleleft c_4$ gives more information than the one directly represented by $c_1 \prec c_4$. The proposed reduction is based on this property.

% % \begin{mydef}
% % The dependency graph $\Gamma_{task}(V,A)$ is reduced to the dependency graph $\Gamma_{task}(V,A')$, where $A' \subseteq A$. A direct arc $(\overset{\frown}{c_i,c_j})$, $i<j$, of $A$, which means that $c_i \prec c_j$, is included in $A'$ if no longer path $c_i \blacktriangleleft c_j$ exists in $A$ because of transitivity.
% % \end{mydef}
% \begin{mydef}
% A DAG is transitive if it contains an arc $(\overset{\frown}{u,v})$ between any two vertices such that there is a path from $u$ to $v$. The transitive closure of a DAG $G=(V,E)$, is the DAG $G_T=(V,E_T)$ for which $E_T$ is the minimal subset of $V \times V$ that includes $E$ and makes $G_T$ transitive.
% \end{mydef}

@@

%------------------------------
% \subsection{Hybrid parallelism}
% It is also possible to combine data and task parallelization techniques to get hybrid parallelism, sometimes more efficient on hybrid architectures. A coarse-grain data parallelism creates for $k$ resources $k$ multi-stencil computations $\mathcal{MSP}_k(T,\mathcal{M}_k,\Delta_k,\Gamma_{data})$, where $\Gamma_{data}$ is built from an ordered list of computations and is itself an ordered list of computations $c_i$ and updated-computations $*_i$. On the other hand, the task parallelization technique builds, from an ordered list of computation, a graph dependency $\Gamma_{task}$. As a result $T_{data}$ and $T_{task}$ can be composed to build $\Gamma_{hybrid}$.

% However, we have to clarify the dependency operator $\prec$ for the specific case of updated-computations. As explained in Section~\ref{sect:dataparal}, if $c_j$ is an updated-computations, which means that at least one $c_i$ exists such that $i<j$ and $c_i \pprec c_j$, $c_j$ is transformed to the ordered list of two computations

@@

% As a result, and because $*_j$ is itself a computation, it appears dependencies relations. First, by definition, the dependency $c^*_j \prec *_j$ is created because $\{w^*_j\} \subset R_j \cup \{w^*_j\}$. Second, as $R^*_j = \{\bigcup_{i} \{w_i\} \text{, }\forall i<j\text{, and }c_i \pprec c_j\}$, for each $i<j$ the dependency $c_i \prec *_j$ is also created.

% The choice of separating $*_j$ from $c_j$ has been done to entirely expose the eventual task parallelism in the hybrid parallelization. For example, Figure~\ref{fig:more} give an exemple where initially $c_0 \pprec c_2$ and $c_0 \prec c_1$, $c_1 \prec c_2$. Thus, $c_2$ is replaced first in $\Gamma_{data}$ by $[*_2(\{w_0\},w^*_2),c^*_2(R_2 \cup \{w^*_2\},\text{exp}_2)]$. As a result, when applying $T_{task}$ on $\Gamma_{data}$, it appears that $*_2$ can be performed in parallel with $c_1$. This is due to the fact that $*_2$ is splitted from from $c_2$ and that $R^*_2=\{w_0\}$ is a subset of $R_2=\{w_0,w_1\}$ and not equal to $R_2$.

@@

% As a result, the hybrid parallelization of a multi-stencil program $\mathcal{MSP}$, produces $k$ multi-stencil programs
% \begin{equation*}
% \mathcal{MSP}_k(T,\mathcal{M}_k,\Delta_k,\Gamma_{hybrid}).
% \end{equation*}
% The set of computations $\Gamma_{hybrid}$ is a dependency graph between computations $c_i$ and update computations $*_i$, and can be built from the call to 
% \begin{equation*}
% T_{task}(T_{data}(\Gamma,0),0).
% \end{equation*}

>>>>

% Multi-stencil mesh-based numerical simulation can be parallelized in various ways and is an interesting kind of application to take advantage of modern heterogeneous HPC architectures, mixing clusters, multi-cores CPUs, vectorization units, GPGPU and many-core accelerators.

@@

\begin{mydef}
\textit{
For $n$ the number of computations in $\Gamma$, and for $i,j$ such that $i<j<n$, a \textit{synchronisation} is needed between $k_i$ and $k_j$, denoted $k_i \pprec k_j$, if $\exists (r,n) \in R_j$ with $n\neq identity$ ($k_j$ is a stencil computation), and such that $\{w_i\} \subset R_j$. Moreover, the data to update is $\{w_i\} \cap R_j = \{w_i$\}.}
\end{mydef}

@@

\begin{mydef}
\textit{A synchronization between two computations $k_i \pprec k_j$ is defined as a specific computation $sync(R,w,exp,D)$ where $R=\{w_i\}$, $w=w_i$, $exp=identity$ and $D=\emptyset$.}
\end{mydef}

\begin{mydef}
If a synchronization is needed before $k_j$ (on the data $w_i$), $k_j$ is replaced in $\Gamma_{sync}$ by the list
\begin{equation*}
\Gamma_{sync} = [sync(\{w_i\},w_i,identity,\emptyset), k_j]
\end{equation*}
\end{mydef}

\begin{mydef}
The concatenation of two ordered lists of respectively $n$ and $m$ computations $l_1=[k_i]_{0 \leq i \leq n-1}$ and $l_2=[c'_i]_{0 \leq i \leq m-1}$ is denoted $l_1 \cdot l_2$ and is equal to a new ordered list $l_3=[c_0,\dots,c_{n-1},c'_0,\dots,c'_{m-1}]$.
\end{mydef}

\begin{mydef}
From the ordered list of computation $\Gamma$, a new ordered list $\Gamma_{sync}$ is obtained from the call $\Gamma_{sync} = F_{sync}(\Gamma,0)$, where $F_{sync}$ is the recursive function defined as
\begin{equation*}
F_{sync}(\Gamma,j) = 
\begin{cases} 	[\Gamma[j]] \cdot F_{sync}(\Gamma,j+1) & \mbox{if }\forall i<j \mbox{, } k_i \not\pprec k_j\\
				[sync(\{w_i\},w_i,identity,\emptyset), k_j] \cdot F_{sync}(\Gamma,j+1) & \forall i<j \text{, and }k_i \pprec k_j\}\\
				[] & \mbox{if }j=|\Gamma|.
\end{cases}
\end{equation*}
\end{mydef}

@@

% \begin{myprop}
% Denoting four computations $c_l$, $c_k$, $k_i$ and $k_j$ of $\Gamma$, with $l<k$, $i<j$, and where $c_l \pprec c_k$ and $k_i \pprec k_j$. Then, computations $c_k$ and $k_j$ are tansformed to $c^*_k(c_k,update(w_l))$ and $c^*_j(k_j,update(w_i))$. If $i<k$, a single update call can be performed at $k$ such that $c^*_k(c_k,update(w_l\cup w_i))$, and in this case, the computation $k_j$ is not transformed.
% \end{myprop}

% \begin{proof}
% By definition, the synchronization of $w_i$ can be performed between $k_i$ and $k_j$. As $i<k$ the computation $k_i$ is not performed between $k$ and $j$. As a result, the synchronization of $w_i$ can be performed between $k_i$ and $c_k$.
% \end{proof}

% In other words, for each synchronization in $\Gamma_{data}$, if the computation $k_i$ from which $k_j$ is synchronized occurs before the last synchronization, the two synchronizations can be reduced to one with the union of the data to update.

@@

\begin{mydef}
For two computations $k_i(R_i,w_i,\text{exp}_i,D_i)$ and $k_j(R_j,w_j,\text{exp}_j,D_j)$, with $i < j$, it is said that $k_j$ is data dependant from $k_i$, denoted $k_i\prec k_j$, if $\{w_i\} \cap R_j \neq \emptyset$. In this case, $k_i$ has to be computed before $k_j$. The binary relation $k_i\prec k_j$ represents a \textit{dependency}.
\end{mydef}

@@

% \begin{mydef}
% For two computations $k_i(R_i,w_i,\text{exp}_i)$ and $k_j(R_j,w_j,\text{exp}_j)$, a time dependency is denoted $k_i \blacktriangleleft k_j$ and means that $k_i$ has to be computed before $k_j$.
% \end{mydef}

@@

% \begin{myprop}
% For two computations $k_i(R_i,w_i,\text{exp}_i)$ and $k_j(R_j,w_j,\text{exp}_j)$ such that $k_i \prec k_j$, $k_i \blacktriangleleft k_j$ is verified.
% \end{myprop}

% \begin{proof}
% By definition $k_i \prec k_j$ means that $\{w_i\}\cap R_j \neq \emptyset$ and that $k_i$ has to computed before $k_j$. %Thus, the relation $k_i \prec k_j$ is stronger than $k_i \blacktriangleleft k_j$.
% \end{proof}

@@

From an ordered list of computations $\Gamma_{sync}$, a directed dependency graph $\Gamma_{dep}(V,A)$ can be built finding all pairs of computations $k_i(R_i,w_i,\text{exp}_i,D_i)$ and $k_j(R_j,w_j,\text{exp}_j,D_j)$, with $i<j$, such that $k_i \prec k_j$. %Those time dependencies can be found from pairs of data dependencies $k_i \prec k_j$.

@@

\begin{mydef}
From the ordered list $\Gamma_{sync}$ of computations $c(R,w,\text{exp},D)$, a directed dependency graph $\Gamma_{dep}(V,A)$ is obtained from the call $T_{dep}(\Gamma_{sync},0)$, where $F_{dep}$ is the recursive function
\begin{equation*}
F_{dep}(\Gamma_{sync},j) = 
\begin{cases} 	(\{\},\{\}) & \mbox{if }j=|\Gamma_{sync}|\\
				(k_j, \{(\overset{\frown}{k_i,k_j})\mbox{, }\forall i < j \mbox{, } k_i\prec k_j \})\cup F_{dep}(\Gamma_{sync},j+1) & \mbox{if }j<|\Gamma_{sync}|
\end{cases}
\end{equation*}
\end{mydef}

This constructive function is possible because the input is an ordered list. Actually, if $k_i\prec k_j$ then $i<j$. As a result, $k_i$ is already in $V$ when the arc $(\overset{\frown}{k_i,k_j})$ is built.

@@

As a result, the hybrid parallelization of a multi-stencil program $\mathcal{MSP}$, produces $k$ multi-stencil programs
\begin{equation*}
\mathcal{MSP}_k(T,\mathcal{M}_k,\mathcal{E},\mathcal{D},\Delta_k,\Gamma_{dep}).
\end{equation*}
The set of computations $\Gamma_{dep}$ is a dependency graph between computations $k_i$ of $\Gamma$ and the synchronization computations $synk_i$ added into $\Gamma_{sync}$. $\Gamma_{dep}$ can be built from the call to 
\begin{equation*}
F_{dep}(F_{sync}(\Gamma,0),0).
\end{equation*}

% Using the function $T_{task}$ to build $\Gamma_{task}$, however, duplication of dependencies may occur because of the transitivity of the relation $\blacktriangleleft$. Actually, as the relation $\prec$ verifies the relation $\blacktriangleleft$, and as $\blacktriangleleft$ is transitive, if for three computations $c_k(R_k,w_k,\text{exp}_k)$, $k_i(R_i,w_i,\text{exp}_i)$ and $k_j(R_j,w_j,\text{exp}_j)$, with $k<i<j$ and where $c_k \prec k_i$, $k_i \prec k_j$ and $c_k \prec k_j$, then the transitivity $c_k \blacktriangleleft k_j$ is verified because of the relations $c_k \blacktriangleleft k_i$ and $k_i \blacktriangleleft k_j$. However, $c_k \blacktriangleleft k_j$ is also directly represented by the binary relation $c_k \prec k_j$. As a result, a duplication of the dependency is created in $\Gamma_{task}$. In Figure~\ref{fig:duplication} an example of duplication is given for the relation $c_1 \blacktriangleleft c_4$.

@@

% Another view of the relations $\prec$ and $\blacktriangleleft$ is that $k_i\blacktriangleleft k_j$ in $\Gamma_{task}$ is a path from $k_i$ to $k_j$ of any size, while $k_i \prec k_j$ is a path of size $1$ from $k_i$ to $k_j$. And in any case a path in the graph represents a time dependency between computations. Scheduling a graph dependency which do not contains duplication of information is easier. 
% %For this reason, we define a reduction to apply to $\Gamma_{task}$, and which is performed as a post-transformation. In the example of Figure~\ref{fig:duplication} the dependency represented by the path $c_1 \blacktriangleleft c_2 \blacktriangleleft c_3 \blacktriangleleft c_4$ gives more information than the one directly represented by $c_1 \prec c_4$. The proposed reduction is based on this property.

% % \begin{mydef}
% % The dependency graph $\Gamma_{task}(V,A)$ is reduced to the dependency graph $\Gamma_{task}(V,A')$, where $A' \subseteq A$. A direct arc $(\overset{\frown}{k_i,k_j})$, $i<j$, of $A$, which means that $k_i \prec k_j$, is included in $A'$ if no longer path $k_i \blacktriangleleft k_j$ exists in $A$ because of transitivity.
% % \end{mydef}
% \begin{mydef}
% A DAG is transitive if it contains an arc $(\overset{\frown}{u,v})$ between any two vertices such that there is a path from $u$ to $v$. The transitive closure of a DAG $G=(V,E)$, is the DAG $G_T=(V,E_T)$ for which $E_T$ is the minimal subset of $V \times V$ that includes $E$ and makes $G_T$ transitive.
% \end{mydef}

@@

%------------------------------
% \subsection{Hybrid parallelism}
% It is also possible to combine data and task parallelization techniques to get hybrid parallelism, sometimes more efficient on hybrid architectures. A coarse-grain data parallelism creates for $k$ resources $k$ multi-stencil computations $\mathcal{MSP}_k(T,\mathcal{M}_k,\Delta_k,\Gamma_{data})$, where $\Gamma_{data}$ is built from an ordered list of computations and is itself an ordered list of computations $k_i$ and updated-computations $*_i$. On the other hand, the task parallelization technique builds, from an ordered list of computation, a graph dependency $\Gamma_{task}$. As a result $T_{data}$ and $T_{task}$ can be composed to build $\Gamma_{hybrid}$.

% However, we have to clarify the dependency operator $\prec$ for the specific case of updated-computations. As explained in Section~\ref{sect:dataparal}, if $k_j$ is an updated-computations, which means that at least one $k_i$ exists such that $i<j$ and $k_i \pprec k_j$, $k_j$ is transformed to the ordered list of two computations

@@

% As a result, and because $*_j$ is itself a computation, it appears dependencies relations. First, by definition, the dependency $c^*_j \prec *_j$ is created because $\{w^*_j\} \subset R_j \cup \{w^*_j\}$. Second, as $R^*_j = \{\bigcup_{i} \{w_i\} \text{, }\forall i<j\text{, and }k_i \pprec k_j\}$, for each $i<j$ the dependency $k_i \prec *_j$ is also created.

% The choice of separating $*_j$ from $k_j$ has been done to entirely expose the eventual task parallelism in the hybrid parallelization. For example, Figure~\ref{fig:more} give an exemple where initially $c_0 \pprec c_2$ and $c_0 \prec c_1$, $c_1 \prec c_2$. Thus, $c_2$ is replaced first in $\Gamma_{data}$ by $[*_2(\{w_0\},w^*_2),c^*_2(R_2 \cup \{w^*_2\},\text{exp}_2)]$. As a result, when applying $T_{task}$ on $\Gamma_{data}$, it appears that $*_2$ can be performed in parallel with $c_1$. This is due to the fact that $*_2$ is splitted from from $c_2$ and that $R^*_2=\{w_0\}$ is a subset of $R_2=\{w_0,w_1\}$ and not equal to $R_2$.

@@

% As a result, the hybrid parallelization of a multi-stencil program $\mathcal{MSP}$, produces $k$ multi-stencil programs
% \begin{equation*}
% \mathcal{MSP}_k(T,\mathcal{M}_k,\Delta_k,\Gamma_{hybrid}).
% \end{equation*}
% The set of computations $\Gamma_{hybrid}$ is a dependency graph between computations $k_i$ and update computations $*_i$, and can be built from the call to 
% \begin{equation*}
% T_{task}(T_{data}(\Gamma,0),0).
% \end{equation*}
